Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-06-02 paper/main.tex
suggestion of a new layout of Table 2

\begin{table}[t]
  \caption{Considered translators and the numbers of \emph{fails} and
    successfully constructed automata with \emph{at most 1 mark} and
    with \emph{at least 2 marks} for each translator and set of
    formulas.}
\label{tab:ltltranslators}
\centering
% \begin{tabular}{lr@{~~~~~~}lr}
% \toprule
% tool        & version&homepage & ~~~automata \\
% \midrule
% \texttt{ltl2dela}~~~ & 21.0 & \multirow{2}{*}{\url{https://owl.model.in.tum.de/}} & 129\\
% \texttt{ltl2dgra} & 21.0 & & 234\\
% \texttt{ltl2tgba} & 2.10.4 & \url{https://spot.lrde.epita.fr} & 70\\
% \texttt{ltl3tela} & 2.2.0  & \url{https://github.com/jurajmajor/ltl3tela} & 91\\
% \bottomrule
% \end{tabular}
%
% \begin{tabular}{lrrrr}
%   \toprule
%   &&& \multicolumn{2}{c}{constructed automata with} \\ \cmidrule(lr){4-5}
%   tool &\!\!\!version/homepage & ~~~~fails & less than 2 marks & \textbf{at least 2 marks}\\
%   \midrule
%   \texttt{ltl2dela} & 21.0\tablefootnote{\label{fn:owl}\url{https://owl.model.in.tum.de/}}
%        & 5 & 214 & \textbf{129} \\
%   \texttt{ltl2dgra} & 21.0\footnotemark[1]
%        & 12 & 102 & \textbf{234}\\
%   \texttt{ltl2tgba} & 2.10.4\tablefootnote{\url{https://spot.lrde.epita.fr}}
%        & 0 & 278 & \textbf{70}\\
%   \texttt{ltl3tela} & 2.2.0\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
%        & 18 & 239 & \textbf{91}\\
%   \bottomrule
% \end{tabular}
%

\setlength{\tabcolsep}{5.7pt}
\begin{tabular}{lrrrrrrr}
  \toprule
  && \multicolumn{3}{c}{\!348 formulas from literature\!} & \multicolumn{3}{c}{366 random formulas}\\
  \cmidrule(lr){3-5} \cmidrule(lr){6-8}
  &&& \multicolumn{2}{c}{automata with} && \multicolumn{2}{c}{automata with}\\
  \cmidrule(lr){4-5} \cmidrule(lr){7-8}
  &&& at most & \textbf{at least} && at most & \textbf{at least}\\[-.8ex]
  \multicolumn{2}{l}{translator ~~~ (version)$^{\text{web}}$} & ~~~~fails & 1 mark & \!\!\!\!\textbf{2 marks} & ~~~~fails & 1 mark & \!\!\!\!\textbf{2 marks}\\
  \midrule 
  \texttt{ltl2tgba -G} & (2.10.4)\tablefootnote{\url{https://spot.lrde.epita.fr}}
  & 0 & 278 & \textbf{70} & ? & ? & \textbf{?}\\
  \texttt{ltl3tela} & (2.2.0)\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
  & 18 & 239 & \textbf{91} & ? & ? & \textbf{?}\\
  \texttt{ltl3tela -D1} & (2.2.0)\footnotemark[2]
  & ? & ? & \textbf{?} & ? & ? & \textbf{?}\\ 
  \texttt{ltl2dela} & (21.0)\tablefootnote{\label{fn:owl}\url{https://owl.model.in.tum.de/}}
       & 5 & 214 & \textbf{129} & ? & ? & \textbf{?} \\
  \texttt{ltl2dgra} & (21.0)\footnotemark[3]
       & 12 & 102 & \textbf{234} & ? & ? & \textbf{?}\\
  \bottomrule
\end{tabular}
\end{table}

\fixme{Honza je tu.}
To each automata set (that is, the automata with at least two marks
produced by an individual LTL translator), we have applied all
single-level reductions and the multi-level reduction, always with
incremental approach and without reducing the number of cubes. We
do not reduce the number of cubes as our primary aim is to reduce the
number of acceptance marks.
%\todo{Vysvetlit, proc neredukujeme pocet cubes?}
The timeout for each QBF query was set to 30 seconds.
%
All reductions have been performed by the tool \texttt{telatko}
built with Spot library version 2.10.4 and Z3 version 4.8.15. The
experiments have been run on a computer with Intel$^\text{®}$ Core™ i7-8700
processor and 32 GB of memory running Ubuntu 20.04.4. We used the tool
\texttt{autcross} of the Spot library to get the statistics of the
reduced automata and the running times.

Table~\ref{tab:cumm-data} shows the cumulative numbers of marks in the
input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
column \textit{solver timeout} shows the number of automata for which
the last query to QBF solver did not finish within the 30~seconds
limit. The timeout of the last QBF query means that the automaton may
be potentially further reduced if a longer time limit is used. One can
observe that a higher level sometimes achieves a smaller reduction
than a lower level (e.g., compare level~1 and level~2 for
\texttt{ltl2dgra}). This is caused by the QBF solver timeouts occuring
earlier as formulas constructed by the higher level are more
complex. For \texttt{ltl2dela} and \texttt{ltl2dgra}, the automata
sets do not contain any automaton where level~2 or level~3 achieves a
better result than level~1. However, all levels contribute to the
reductions in the multi-level setting.

\begin{table}[t]
  \caption{The cumulative numbers of acceptance marks before and after
    reduction for various reductions and automata sets. The column
    \emph{reduction} shows the cumulative percentage of saved
    acceptance marks and \emph{time} reports the cumulative reduction
    time. The column \textit{solver timeout} indicates the number of
    instances where the last call to the QBF solver timed out.  }
\label{tab:cumm-data}
\centering
\setlength{\tabcolsep}{10pt}
\input{table-data.tex}
\end{table}

Table~\ref{tab:heatmap} shows the number of automata from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2~marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 24 cases, we have reduced 7 or more
acceptance marks to only 4.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.

>>>>

\begin{table}[t]
  \caption{Considered translators and the numbers of \emph{fails} and
    successfully constructed automata with \emph{at most 1 mark} and
    with \emph{at least 2 marks} for each translator and set of
    formulas.}
\label{tab:ltltranslators}
\centering
% \begin{tabular}{lr@{~~~~~~}lr}
% \toprule
% tool        & version&homepage & ~~~automata \\
% \midrule
% \texttt{ltl2dela}~~~ & 21.0 & \multirow{2}{*}{\url{https://owl.model.in.tum.de/}} & 129\\
% \texttt{ltl2dgra} & 21.0 & & 234\\
% \texttt{ltl2tgba} & 2.10.4 & \url{https://spot.lrde.epita.fr} & 70\\
% \texttt{ltl3tela} & 2.2.0  & \url{https://github.com/jurajmajor/ltl3tela} & 91\\
% \bottomrule
% \end{tabular}
%
% \begin{tabular}{lrrrr}
%   \toprule
%   &&& \multicolumn{2}{c}{constructed automata with} \\ \cmidrule(lr){4-5}
%   tool &\!\!\!version/homepage & ~~~~fails & less than 2 marks & \textbf{at least 2 marks}\\
%   \midrule
%   \texttt{ltl2dela} & 21.0\tablefootnote{\label{fn:owl}\url{https://owl.model.in.tum.de/}}
%        & 5 & 214 & \textbf{129} \\
%   \texttt{ltl2dgra} & 21.0\footnotemark[1]
%        & 12 & 102 & \textbf{234}\\
%   \texttt{ltl2tgba} & 2.10.4\tablefootnote{\url{https://spot.lrde.epita.fr}}
%        & 0 & 278 & \textbf{70}\\
%   \texttt{ltl3tela} & 2.2.0\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
%        & 18 & 239 & \textbf{91}\\
%   \bottomrule
% \end{tabular}
%
\setlength{\tabcolsep}{5.7pt}
\begin{tabular}{lrrrrrrr}
  \toprule
  && \multicolumn{3}{c}{\!348 formulas from literature\!} & \multicolumn{3}{c}{366 random formulas}\\
  \cmidrule(lr){3-5} \cmidrule(lr){6-8}
  &&& \multicolumn{2}{c}{automata with} && \multicolumn{2}{c}{automata with}\\
  \cmidrule(lr){4-5} \cmidrule(lr){7-8}
  &&& at most & \textbf{at least} && at most & \textbf{at least}\\[-.8ex]
  \multicolumn{2}{l}{translator ~~~ (version)$^{\text{web}}$} & ~~~~fails & 1 mark & \!\!\!\!\textbf{2 marks} & ~~~~fails & 1 mark & \!\!\!\!\textbf{2 marks}\\
  \midrule 
  \texttt{ltl2tgba -G} & (2.10.4)\tablefootnote{\url{https://spot.lrde.epita.fr}}
  & 0 & 278 & \textbf{70} & ? & ? & \textbf{?}\\
  \texttt{ltl3tela} & (2.2.0)\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
  & 18 & 239 & \textbf{91} & ? & ? & \textbf{?}\\
  \texttt{ltl3tela -D1} & (2.2.0)\footnotemark[2]
  & ? & ? & \textbf{?} & ? & ? & \textbf{?}\\ 
  \texttt{ltl2dela} & (21.0)\tablefootnote{\label{fn:owl}\url{https://owl.model.in.tum.de/}}
       & 5 & 214 & \textbf{129} & ? & ? & \textbf{?} \\
  \texttt{ltl2dgra} & (21.0)\footnotemark[3]
       & 12 & 102 & \textbf{234} & ? & ? & \textbf{?}\\
  \bottomrule
\end{tabular}
\end{table}

To each automata set (that is, the automata with at least two marks
produced by an individual LTL translator for one of the two formula
sets), we have applied all single-level reductions and the multi-level
reduction, always with incremental approach and without reducing the
number of cubes. We do not reduce the number of cubes as our primary
aim is to reduce the number of acceptance marks.
%\todo{Vysvetlit, proc neredukujeme pocet cubes?}
The timeout for each QBF query was set to 30 seconds.
%
All reductions have been performed by the tool \texttt{telatko}
built with Spot library version 2.10.4 and Z3 version 4.8.15. The
experiments have been run on a computer with Intel$^\text{®}$ Core™ i7-8700
processor and 32 GB of memory running Ubuntu 20.04.4. We used the tool
\texttt{autcross} of the Spot library to get the statistics of the
reduced automata and the running times.

For each automata set identified by the translator and the set of
formulas, Table~\ref{tab:cumm-data} shows the cumulative numbers of
marks in the input automata sets and after each reduction, together
with the reduction ratio and total time spent by the considered
reduction. The column \textit{solver timeout} shows the number of
automata for which the last query to QBF solver did not finish within
the 30~seconds limit. The timeout of the last QBF query means that the
automaton may be potentially further reduced if a longer time limit is
used. \fixme{Zkontrolovat zbytek odstavce podle aktualnich vysledku,
  specifikovat formula set.} One can observe that a higher level
sometimes achieves a smaller reduction than a lower level (e.g.,
compare level~1 and level~2 for \texttt{ltl2dgra}). This is caused by
the QBF solver timeouts occuring earlier as formulas constructed by
the higher level are more complex. For \texttt{ltl2dela} and
\texttt{ltl2dgra}, the automata sets do not contain any automaton
where level~2 or level~3 achieves a better result than
level~1. However, all levels contribute to the reductions in the
multi-level setting.

\begin{table}[t]
  \caption{The cumulative numbers of acceptance marks before and after
    reduction for various reduction levels and automata sets given by
    the translator and formula set. The column \emph{reduction} shows
    the cumulative percentage of saved acceptance marks and
    \emph{time} reports the cumulative reduction time. The column
    \textit{solver timeout} indicates the number of instances where
    the last call to the QBF solver timed out.  }
\label{tab:cumm-data}
\centering
\setlength{\tabcolsep}{3.9pt}
\newcommand{\mybox}[1]{\fbox{\begin{minipage}{4.7cm}\centering#1\end{minipage}}}
\begin{tabular}{lcrrrrrrrr}
  \toprule
  && \multicolumn{4}{c}{\makecell{reduction of automata from\\[-.8ex] formulas from literature}}
  & \multicolumn{4}{c}{\makecell{reduction of automata from\\[-.8ex] random formulas}}
  \\ \cmidrule(lr){3-6} \cmidrule(lr){7-10}
  translator & \makecell{reduction\\[-1ex] level}
  & marks & \makecell{reduction\\[-1ex] [\%]} & \makecell{time\\[-1ex] [s]} & \makecell{solver\\[-1ex] timeout} 
  & marks & \makecell{reduction\\[-1ex] [\%]} & \makecell{time\\[-1ex] [s]} & \makecell{solver\\[-1ex] timeout} \\
  \midrule
%  \multirow{5}{*}{\makecell{\texttt{ltl2tgba -G}}} &
  \multicolumn{2}{l}{\texttt{ltl2tgba -G}} 
  & \multicolumn{4}{c}{\mybox{198 marks in 70 automata}}
  & \multicolumn{4}{c}{\mybox{XYZ marks in XYZ automata}}\\
  & 1
  & 198 & 0.0~~~ & 49.3 & 0~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                     
  & 2
  & 198 & 0.0~~~ & 78.9 & 0~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                     
  & 3
  & 189 & 4.5~~~ & 1222.8 & 7~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                     
  & multi
  & 189 & 4.5~~~ & 1259.1 & 6~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                    
  \midrule
%  \multirow{5}{*}{\makecell{\texttt{ltl2tgba -G}}} &
  \multicolumn{2}{l}{\texttt{ltl3tela}} 
  & \multicolumn{4}{c}{\mybox{348 marks in 91 automata}}
  & \multicolumn{4}{c}{\mybox{XYZ marks in XYZ automata}}\\
  & 1
  & 332 & 4.6~~~ & 465.9 & 12~~~
  &   ? &   ?~~~ &     ? &  ?~~~ \\                     
  & 2
  & 334 & 4.0~~~ & 553.9 & 15~~~
  &   ? &   ?~~~ &     ? &  ?~~~ \\                     
  & 3
  & 326 & 6.3~~~ & 691.8 & 18~~~
  &   ? &   ?~~~ &     ? &  ?~~~ \\                     
  & multi
  & 323 & 7.2~~~ & 1545.9 & 18~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                    
  \midrule
  &&&& \vdots \\
% \\
% level 1 &
% 387~ & 26.0~~~~~~ & 755.4~ &  18~~~
% \\
% &
% level 2 &
% 390~ & 25.4~~~~~~ & 1024.5~ &  20~~~
% \\
% &
% level 3 &
% 393~ & 24.9~~~~~~ & 7273.3~ &  25~~~
% \\
% &
% multi &
% 371~ & 29.1~~~~~~ & 9111.4~ &  21~~~
% \\
% \midrule
% \multirow{4}{*}{\makecell{\texttt{ltl2dgra}\\234 automata\\882 acc.~marks}}
% &
% level 1 &
% 552~ & 37.4~~~~~~ & 738.7~ & 14~~~
% \\
% &
% level 2 &
% 569~ & 35.5~~~~~~ & 907.3~ &  18~~~
% \\
% &
% level 3 &
% 561~ & 36.4~~~~~~ & 1163.4~ &  21~~~
% \\
% &
% multi &
% 542~ & 38.5~~~~~~ & 2320.7~ &  21~~~
% \\
% \midrule
% \multirow{4}{*}{\makecell{\texttt{ltl3tela}\\91 automata\\348 acc.~marks}}
% &
% level 1 &
% 332~ & 4.6~~~~~~ & 465.9~ &  12~~~
% \\
% &
% level 2 &
% 334~ & 4.0~~~~~~ & 553.9~ &  15~~~
% \\
% &
% level 3 &
% 326~ & 6.3~~~~~~ & 691.8~ &  18~~~
% \\
% &
% multi &
% 323~ & 7.2~~~~~~ & 1545.9~ & 18~~~
% \\
\bottomrule                                                                     
\end{tabular}
\end{table}

\fixme{Upravit zbytek textu.}
Table~\ref{tab:heatmap} shows the number of automata from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2~marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 24 cases, we have reduced 7 or more
acceptance marks to only 4.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.
